Supplementary Material for the article \textit{Quotient Normalized Maximum Likelihood Criterion for Learning Bayesian Network Structures}

### 2018-02-14 bn.tex
More minor typos

Score-based Bayesian learning of Bayesian network structures evaluates the goodness of different structures $G$ using their
posterior probability $P(G|D,\alpha)$, where $\alpha$ denotes the
hyperparameters for the model parameters $\theta$, and $D$ is a
collection of $N$ $n$\nobreakdash-dimensional i.i.d. data vectors
collected to a $N\times n$ design matrix. We use the notation $D_i$ to
denote the $i^\text{th}$ column of the data matrix and the notation $D_V$ to denote
the columns that correspond to the variable subset $V$. We also write
$D_{i,G_i}$ for $D_{\{i\}\cup G_i}$ and denote the entries of the
column $i$ on the rows on which the parents $G_i$ contain the value
configuration number by $D_{i,G_i=j}$, $j\in\{1,\ldots,q_i\}$.

@@

Recently, Suzuki~\cite{Suzuki2017} discussed the theoretical
properties of the BDeu score and showed that in certain settings BDeu
has the tendency to add more and more parent variables for a child node
even though the empirical conditional entropy of the child given the
parents has already reached zero. In more detail, assume that in our
data $D$ the values of $X_i$ are completely determined by variables
in set $Z$, so that the empirical entropy $H_N(X_i | Z)$ is zero. Now, if we can further
find one or more variables, denoted by $Y$, whose values are
determined completely by the variables in $Z$, then BDeu will prefer
the set $Z\cup Y$ over $Z$ alone as the parents of $X_i$. Suzuki
argues that this kind of behavior violates \textit{regularity} in model
selection as the more complex model is preferred over a simpler one
even though it does not fit the data any better. The phenomenon seems
to stem from the way the hyperparameters for the Dirichlet
distribution are chosen in BDeu as using Jeffreys' prior,
$\theta_{ijk}\sim Dir(\frac{1}{2},\ldots,\frac{1}{2})$, does not suffer
from this anomaly. However, using Jeffreys' prior causes marginal
likelihood score not to be score equivalent. In Section \ref{sec:regularity}, we will give the formal definition of regularity and state that qNML is regular. In addition, we provide a proof of regularity for fNML criterion, which has not appeared in the literature before. The detailed proofs can be found in Appendix B in the Supplementary Material.

@@

Consequently, Silander et al.~\cite{cosco.pgm08a}
suggested solving the BDeu parameter sensitivity problem by using the
NML code for the column partitions, i.e., changing the Bayesian mixture
in equation~(\ref{eqn:bayesmix}) to
\begin{equation}
P^1_{NML}(D_{i,G_i=j};G)=\frac{P(D|\hat\theta(D_{i,G_i=j};G))}{\sum_{D'}{P(D'|\hat\theta(D';G))}},
\end{equation}
where $D'\in{\{1,\ldots,r_i\}}^{|D_{i,G_i=j}|}$.  The logarithm of the
denominator is often called a regret, since it indicates the extra
code length needed compared to the code length obtained using the (a
priori unknown) maximum likelihood parameters. The regret for
$P^1_{NML}$ depends only on the length $N$ of the categorical data
vector with $r$ different categorical values,
\begin{equation}
reg(N,r)=\log \sum_{D\in \{1,\ldots,r\}^N} P(D|\hat\theta(D)).
\end{equation}
While the naive
computation of the regret is still prohibitive, Silander et al.\
approximate it efficiently using a so-called Szpankowski
approximation~\cite{cosco.aistat03}:
\begin{eqnarray}
\label{eqn:szp1}
\lefteqn{reg(N,r) \approx \frac{\sqrt{2} r \Gamma{\left(\frac{r}{2} \right)}}
                               {3 \sqrt{N} \Gamma{\left(\frac{r-1}{2}  \right)}}} \\
&&+ \left(\frac{r-1}{2}\right) \log{\left (\frac{N}{2} \right )}
- \log \Gamma{\left(\frac{r}{2} \right)} + \frac{1}{2} \log{\left (\pi \right )}\nonumber\\
&&- \frac{r^{2} \Gamma^{2}{\left(\frac{r }{2} \right)}}
         {9N \Gamma^{2}{\left(\frac{r-1}{2}\right)}}
+ \frac{2r^3-3r^2-2r+3}{36N}\nonumber.
\end{eqnarray}

>>>>

Score-based Bayesian learning of Bayesian network structures evaluates the goodness of different structures $G$ using their
posterior probability $P(G|D,\alpha)$, where $\alpha$ denotes the
hyperparameters for the model parameters $\theta$, and $D$ is a
collection of $N$ $n$\nobreakdash-dimensional i.i.d. data vectors
collected to a $N\times n$ design matrix. We use the notation $D_i$ to
denote the $i^\text{th}$ column of the data matrix and the notation $D_V$ to denote
the columns that correspond to the variable subset $V$. We also write
$D_{i,G_i}$ for $D_{\{i\}\cup G_i}$ and denote the entries of the
column $i$ on the rows on which the parents $G_i$ contain the value
configuration number $j$ by $D_{i,G_i=j}$, $j\in\{1,\ldots,q_i\}$.

@@

Recently, Suzuki~\cite{Suzuki2017} discussed the theoretical
properties of the BDeu score and showed that in certain settings BDeu
has the tendency to add more and more parent variables for a child node
even though the empirical conditional entropy of the child given the
parents has already reached zero. In more detail, assume that in our
data $D$ the values of $X_i$ are completely determined by variables
in a set $Z$, so that the empirical entropy $H_N(X_i | Z)$ is zero. Now, if we can further
find one or more variables, denoted by $Y$, whose values are
determined completely by the variables in $Z$, then BDeu will prefer
the set $Z\cup Y$ over $Z$ alone as the parents of $X_i$. Suzuki
argues that this kind of behavior violates \textit{regularity} in model
selection as the more complex model is preferred over a simpler one
even though it does not fit the data any better. The phenomenon seems
to stem from the way the hyperparameters for the Dirichlet
distribution are chosen in BDeu as using Jeffreys' prior,
$\theta_{ijk}\sim Dir(\frac{1}{2},\ldots,\frac{1}{2})$, does not suffer
from this anomaly. However, using Jeffreys' prior causes marginal
likelihood score not to be score equivalent. In Section \ref{sec:regularity}, we will give the formal definition of regularity and state that qNML is regular. In addition, we provide a proof of regularity for fNML criterion, which has not appeared in the literature before. The detailed proofs can be found in Appendix B in the Supplementary Material.

@@

Consequently, Silander et al.~\cite{cosco.pgm08a}
suggested solving the BDeu parameter sensitivity problem by using the
NML code for the column partitions, i.e., changing the Bayesian mixture
in equation~(\ref{eqn:bayesmix}) to
\begin{equation}
P^1_{NML}(D_{i,G_i=j};G)=\frac{P(D|\hat\theta(D_{i,G_i=j};G))}{\sum_{D'}{P(D'|\hat\theta(D';G))}},
\end{equation}
where $D'\in{\{1,\ldots,r_i\}}^{|D_{i,G_i=j}|}$.  The logarithm of the
denominator is often called the regret, since it indicates the extra
code length needed compared to the code length obtained using the (a
priori unknown) maximum likelihood parameters. The regret for
$P^1_{NML}$ depends only on the length $N$ of the categorical data
vector with $r$ different categorical values,
\begin{equation}
reg(N,r)=\log \sum_{D\in \{1,\ldots,r\}^N} P(D|\hat\theta(D)).
\end{equation}
While the naive
computation of the regret is still prohibitive, Silander et al.\
approximate it efficiently using a so-called Szpankowski
approximation~\cite{cosco.aistat03}:
\begin{eqnarray}
\label{eqn:szp1}
\lefteqn{reg(N,r) \approx \frac{\sqrt{2} r \Gamma{\left(\frac{r}{2} \right)}}
                               {3 \sqrt{N} \Gamma{\left(\frac{r-1}{2}  \right)}}} \\
&&+ \left(\frac{r-1}{2}\right) \log{\left (\frac{N}{2} \right )}
- \log \Gamma{\left(\frac{r}{2} \right)} + \frac{1}{2} \log{\left (\pi \right )}\nonumber\\
&&- \frac{r^{2} \Gamma^{2}{\left(\frac{r }{2} \right)}}
         {9N \Gamma^{2}{\left(\frac{r-1}{2}\right)}}
+ \frac{2r^3-3r^2-2r+3}{36N}\nonumber.
\end{eqnarray}

### 2018-02-14 experiments.tex
More minor typos

To empirically compare the model selection criteria we took 20 UCI
data sets~\cite{Lichman:2013} and ran train and test experiments for
all of them. To better compare the performance over different sample
sizes, we picked different fractions ($10\%, 20\%, \ldots, 90\%$) of
the data sets as training data and the rest as the test data.
This was done for 1000 different permutations of each data set.
The training was conducted using the dynamic
programming based exact structure learning algorithm.

@@

For each train/test sample, we ranked the predictive performance of
the models learned by four different structure learning criteria (rank
1 being the best and 4 the worst). Table~\ref{tbl:preds} features the
average rank for different data sets, the average being taken over
1000 different train test samples for each 9 sample sizes for each
data.  BIC's bias for simplicity makes it often win (written bold in
the table) with small sample sizes, but it often performs worst
(underlined) for the larger sample sizes (for the same reason), while
fNML seems to be good for large sample sizes. The striking feature
about the qNML is its robustness.  It is usually between BIC and fNML
for all the sample sizes making it a ``safe choice''. This can be
quantified if we further average the columns of the
table~\ref{tbl:preds}.  This yields the average ranks of $2.95, 2.57,
2.10$, and $2.37$, with standard deviations $0.49, 0.90, 0.76$, and
$0.43$.  While fNML achieves on average the best rank, the runner-up
qNML has the lowest standard deviation.

Figure~\ref{fig:bcnpmean} shows how fNML still sometimes behaves
strangely in terms of model complexity here measured by the number of
parameters in the model. qNML, instead, appears to yield more
parsimonious models. To study the concern of fNML producing too
complex models for small sample sizes, we studied the number of
parameters in models produced by different model selection criteria
when using 10\% of each data set for structure learning.

>>>>

To empirically compare the model selection criteria, we took 20 UCI
data sets~\cite{Lichman:2013} and ran train and test experiments for
all of them. To better compare the performance over different sample
sizes, we picked different fractions ($10\%, 20\%, \ldots, 90\%$) of
the data sets as training data and the rest as the test data.
This was done for 1000 different permutations of each data set.
The training was conducted using the dynamic
programming based exact structure learning algorithm.

@@

For each train/test sample, we ranked the predictive performance of
the models learned by four different structure learning criteria (rank
1 being the best and 4 the worst). Table~\ref{tbl:preds} features the
average rank for different data sets, the average being taken over
1000 different train test samples for each 9 sample sizes for each
data.  BIC's bias for simplicity makes it often win (written bold in
the table) with small sample sizes, but it often performs worst
(underlined) for the larger sample sizes (for the same reason), while
fNML seems to be good for large sample sizes. The striking feature
about the qNML is its robustness.  It is usually between BIC and fNML
for all the sample sizes making it a ``safe choice''. This can be
quantified if we further average the columns of
Table~\ref{tbl:preds}.  This yields the average ranks of $2.95, 2.57,
2.10$, and $2.37$, with standard deviations $0.49, 0.90, 0.76$, and
$0.43$.  While fNML achieves on average the best rank, the runner-up
qNML has the lowest standard deviation.

Figure~\ref{fig:bcnpmean} shows how fNML still sometimes behaves
strangely in terms of model complexity, here measured by the number of
parameters in the model. qNML, instead, appears to yield more
parsimonious models. To study the concern of fNML producing too
complex models for small sample sizes, we studied the number of
parameters in models produced by different model selection criteria
when using 10\% of each data set for structure learning.

### 2018-02-14 qnml.tex
More minor typos

Suzuki \cite{Suzuki2017} independently introduces a Bayesian Dirichlet
quotient (BDq) score that can also be shown to be score equivalent
and regular.  However, like BDeu, this score features a single
hyperparameter $\alpha$, and our initial studies suggest that BDq is
also very sensitive to this hyperparameter (see Appendix D in the
Supplementarty Material), the issue that was one of the main
motivations to develop a parameter-free model selection criterion like
qNML.

>>>>

Suzuki \cite{Suzuki2017} independently introduces a Bayesian Dirichlet
quotient (BDq) score that can also be shown to be score equivalent
and regular.  However, like BDeu, this score features a single
hyperparameter $\alpha$, and our initial studies suggest that BDq is
also very sensitive to this hyperparameter (see Appendix D in the
Supplementary Material), the issue that was one of the main
motivations to develop a parameter-free model selection criterion like
qNML.
